Empowering Domain Recursion in Symmetric Weighted First-Order Model Counting

### 2022-01-23 notes/notes.tex
Support for domain recursion on any domain

\DeclareMathOperator{\Compile}{\textsc{Compile}}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\Imm}{Im}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\WMC}{WMC}

@@

\paragraph{Things I might need to explain.}
\begin{itemize}
\item notation: $\Imm$
\item atom, literal, constant, predicate, variable
\item inequality constraint. Notation: $(x, y)$, where $x$ is a variable (of the clause), and $y$ is either a variable or a constant.
\item $\Vars$, $\Vars(c) = \Vars(L) \cup \Vars(C)$
\item $\Doms$ on both formulas and clauses. $\Doms(c) = \Imm \delta_c$, and $\Doms(\phi) = \bigcup_{c \in \phi} \Doms(c)$.
\item substitution
\item size of a domain, how each domain is partitioned into two during compilation.
\item maybe: notation for partial function, notation for powerset, domain size
\item notation for projection (or avoid it?)
\item $\WMC$
\item constraint removal operation
\item two parts: compilation and inference.
\item introduce and use arrows for bijections, injections, set inclusions, etc.
\item $\emptyset$ for an empty partial map.
\end{itemize}

@@

\paragraph{TODO}
\begin{itemize}
\item later: maybe mathcal instead of mathscr
\item maybe $\pi$ is global enough to have a more unique name
\item maybe $c[\beta, \gamma]$ is better than $c \ast (\beta, \gamma)$
\item avoid using lowercase latin letters for everything: capitalise domains and variables.
\end{itemize}

@@

This $\rho$ is saved as a parameter of the new circuit node. During inference, there is a domain size map $\sigma\colon \mathscr{D} \to \mathbb{N}_0$, which gets updated to
\[
\sigma'(x) =
\begin{cases}
  \sigma(\rho(x)) & \text{if } x \in \dom(\rho) \\
  \sigma(x) & \text{otherwise}
\end{cases}
\]
for all $x \in \mathscr{D}$.

@@

TODO: compilation rules, node types, their evaluation during inference.

@@

\subsection{Improved Domain Recursion}

The original version of domain recursion is here \cite{DBLP:conf/nips/Broeck11}.

@@

TODO: algebraic/graphical representation, what info they hold, and how they're created

@@

>>>>

\DeclareMathOperator{\Compile}{\textsc{Compile}}
\DeclareMathOperator{\CR}{\textsc{CR}}
\DeclareMathOperator{\DR}{\textsc{DR}}
\DeclareMathOperator{\Reff}{\textsc{Ref}}

\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\Imm}{Im}
\DeclareMathOperator{\Doms}{Doms}
\DeclareMathOperator{\size}{size}
\DeclareMathOperator{\Vars}{Vars}
\DeclareMathOperator{\WMC}{WMC}

@@

\paragraph{Things I might need to explain.}
\begin{itemize}
\item notation: $\Imm$
\item atom, (positive/negative) literal, constant, predicate, variable, literal variable
\item inequality constraint. Notation: $(x, y)$, where $x$ is a variable (of the clause), and $y$ is either a variable or a constant.
\item $\Vars$, $\Vars(c) = \Vars(L) \cup \Vars(C)$
\item $\Doms$ on both formulas and clauses. $\Doms(c) = \Imm \delta_c$, and $\Doms(\phi) = \bigcup_{c \in \phi} \Doms(c)$.
\item size of a domain, how each domain is partitioned into two during compilation.
\item maybe: notation for partial function, notation for powerset, domain size
\item notation for projection (or avoid it?)
\item $\WMC$
\item constraint removal operation
\item two parts: compilation and inference.
\item introduce and use arrows for bijections, injections, set inclusions, etc.
\item $\emptyset$ for an empty partial map.
\item notation for variable substitution for literal and constraint sets: $S[x/V]$, where $S$ is the set, $V$ is a set of variable, and $x$ is the new constant or variable. Note that for constraint sets this could mean that a variable and a constant need to switch places.
\item During inference, there is a domain size map $\size\colon \mathscr{D} \to \mathbb{N}_0$.
\end{itemize}

@@

\paragraph{TODO}
\begin{itemize}
\item later: maybe mathcal instead of mathscr
\item maybe $\pi$ is global enough to have a more unique name
\item maybe $c[\beta, \gamma]$ is better than $c \ast (\beta, \gamma)$
\item capitalise variable names.
\end{itemize}

@@

$\WMC(\Reff_\rho(n), \size) = \WMC(n, \size')$ ($n$ is the target circuit node), where $\size'$ is defined as
\[
\size'(x) =
\begin{cases}
  \size(\rho(x)) & \text{if } x \in \dom(\rho) \\
  \size(x) & \text{otherwise}
\end{cases}
\]
for all $x \in \mathscr{D}$.

@@

@@

\paragraph{Evaluation.} $\WMC(\CR_{d \mapsto d'}(n), \size) = \WMC(n, \size \cup \{\, d' \mapsto \size(d) - 1 \,\})$.

TODO: add an explanation for why the way constraints, domain recursion, and constraint removal interact ensures that domain sizes will never be negative.

\subsection{A Generalisation of Domain Recursion}

\begin{algorithm}[t]
  \caption{Formula transformation for domain recursion}
  \label{alg:domainrecursion}
  \KwIn{formula $\phi$, domain $d \in \mathscr{D}$, and constant $x$}
  \KwOut{formula $\phi'$}
  $\phi' \gets \emptyset$\;
  \ForEach{clause $c = (L, C, \delta) \in \phi$\label{line:forclause}}{
    $V \gets \{\, v \in \Vars(L) \mid \delta(v) = d \,\}$\;
    \ForEach{subset $W \subseteq V$ such that $W^2 \cap C = \emptyset$ {\bf and} $W \cap \{\, v \in \Vars(C) \mid (v, y) \in C \text{ for some constant } y \,\} = \emptyset$\label{line:conditions}}{
      $\phi' \gets \phi' \cup \{\, (L[x/W], C[x/W] \cup \{\, (v, x) \mid (v \in V \setminus W) \,\}, \delta') \,\}$\;
      \tcc{$\delta'$ is the restriction of $\delta$ to the new set of variables}
    }
  }
\end{algorithm}

TODO: Compare with the original. The original version of domain recursion is here \cite{DBLP:conf/nips/Broeck11}.

TODO: introduce $\phi$.

\paragraph{Precondition.} Domain recursion can be applied to domain $d \in \mathscr{D}$ if there is a clause $c$ with a literal variable $v \in \Vars(L_c)$ such that $\delta_c(v) = d$.

\paragraph{Operator.} First, introduce a new constant $x$. Then $\DR(\phi) = \DR_d(\Compile(\phi'))$, where $\phi'$ is generated by \cref{alg:domainrecursion}.

\begin{example}
  Let $\phi = \{\, c_1, c_2 \,\}$ be a formula, where
  \begin{align*}
    c_1 &= (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Z, Y) \,\}, \{ X \mapsto a, Y \mapsto b, Z \mapsto b \}), \\
    c_2 &= (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X) \,\}, \{ X \mapsto a, Y \mapsto b, Z \mapsto a \}).
  \end{align*}
  While domain recursion is possible on both domains, here we illustrate how it works on $a$.

  Suppose \cref{line:forclause} picks $c = c_1$ first. Then $V = \{\, X \,\}$. Both subsets of $V$ satisfy the conditions on \cref{line:conditions} and generate new clauses
  \[
  (\{\, \neg p(X, Y), \neg p(X, Z) \,\}, \{\, (Z, Y), (X, x) \,\}, \{ X \mapsto a, Y \mapsto b, Z \mapsto b \}),
  \]
  (from $W = \emptyset$) and
  \[
  (\{\, \neg p(x, Y), \neg p(x, Z) \,\}, \{\, (Z, Y) \,\}, \{ Y \mapsto b, Z \mapsto b \})
  \]
  (from $W = V$).

  When $c = c_2$, then $V = \{\, X, Z \,\}$. The subset $W = V$ fails to satisfy the first condition because of the $Z \ne X$ constraint; without this condition, the resulting clause would have an unsatisfiable constraint $x \ne x$. The other three subsets of $V$ all generate clauses for $\phi'$:
  \[
  (\{\, \neg p(X, Y), \neg p(Z, Y) \,\}, \{\, (Z, X), (X, x), (Z, x) \,\}, \{ X \mapsto a, Y \mapsto b, Z \mapsto a \})
  \]
  (from $W = \emptyset$),
  \[
  (\{\, \neg p(x, Y), \neg p(Z, Y) \,\}, \{\, (Z, x) \,\}, \{ Y \mapsto b, Z \mapsto a \})
  \]
  (from $W = \{\, X \,\}$), and
  \[
  (\{\, \neg p(X, Y), \neg p(x, Y) \,\}, \{\, (X, x) \,\}, \{ X \mapsto a, Y \mapsto b, \})
  \]
  (from $W = \{\, Z \,\}$).
\end{example}

\paragraph{Evaluation.}
\[
\WMC(\DR_d(n), \size) =
\begin{cases}
  \WMC(n, \size) & \text{if } \size(d) > 0 \\
  1 & \text{otherwise.}
\end{cases}
\]

One is picked as the multiplicative identity.

@@

TODO: algebraic/graphical representations

@@

\section{Examples}

\subsection{Counting Partial Injections}
